A Step-Indexed Logic in Agda

### 2023-07-26 papers/July2023/paper.tex
check

This paper fills that need by showing how to build a step-indexed
logic in Agda, a system that we abbreviate as SIL.  This paper is a
literate Agda file; all of the definitions and proofs have been
checked with Agda 2.6.2.2. 
%
Most of the challenge in defining a step-indexed logic is in the
definition of recursive predicates and their fixpoint theorem. The
step-indexed models of \citet{Appel:2001aa} rely on an approximation
operator, here written as $↓_n$, to define the semantic criteria for
valid recursive definitions.  For example, they define a nonexpansive
function to be one that satisfies
\[
  ↓_n f(x) = ↓_n f(↓_n x)
\]
The step-indexed models of \citet{JUNG:2018aa} instead rely on
a step-indexed notion of equality. For example, they define a
nonexpansive function as one that satisfies
\[
  x \stackrel{n}{=} y \text{ implies } f(x) \stackrel{n}{=} f(y)
\]
We are not aware of any inherent reason to prefer one approach to the
other, so we choose the approach of \citet{Appel:2001aa} because they
provide perspicuous proofs of their fixpoint theorem.  So our approach
to SIL can be seen as an implementation of the LSLR logic of
\citet{Dreyer:2011wl} using the foundations of
\citet{Appel:2001aa}. In particular, SIL defines the recursive
predicate of SIL using function iteration, like this:
\[
  ⟦ μ F ⟧ \, x \, n = F^{n\plus 1}\, \top \, x \, n
\]
whereas LSLR uses substitution in the semantics of recursive types:
\[
   ⟦ μ r. R ⟧ \, x \, n = ⟦ R[μr.R/r] ⟧\,x\,n
\]

@@

We organize this paper as follows. We review Appel an McAllester's
work in Section~\ref{sec:appel-mcallester}, then introduce our
representation of step-indexed propositions and predicates in Agda in
Sections \ref{sec:propositions} and \ref{sec:predicates}.  We
formalize several important concepts in
Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
iteration, and prove some basic results about them.  We construct all
of the logical connectives of SIL in
Section~\ref{sec:open-propositions}, including the definition of
recursive predicates and their fixpoint theorem.  The proof theory of
SIL is developed in Section~\ref{sec:proofs}, where we construct the
proof rules of this modal logic.  To evaluate the utility of SIL,
Section~\ref{sec:STLC} presents a case study of proving semantic type
safety for the Simply Typed Lambda Calculus extended with recursive
functions. We find that SIL streamlines the definition of logical
relations and is helpful in many of the proofs, but there is room for
improvement regarding the ergonomics of interactive proof development
using SIL.  Limitations in Agda's inference algorithm sometimes leads
to verbosity in SIL proofs.

\input{latex/July2024/STLC.tex}
\input{latex/July2024/STLCBind.tex}
\input{latex/July2024/STLCSafe.tex}

@@

\input{latex/July2024/StepIndexedLogic.tex}

@@

\input{latex/July2024/STLCDeterministic.tex}

>>>>

This paper fills that need by showing how to build a step-indexed
logic in Agda, a system that we abbreviate as SIL.  This paper is a
literate Agda file; all of the definitions and proofs have been
checked with Agda 2.6.2.2.

UNDER CONSTRUCTION

To evaluate the utility of SIL, Section~\ref{sec:STLC} presents a case
study of proving semantic type safety for the Simply Typed Lambda
Calculus extended with recursive functions. We find that SIL
streamlines the definition of logical relations and is helpful in many
of the proofs, but there is room for improvement regarding the
ergonomics of interactive proof development using SIL.  Limitations in
Agda's inference algorithm sometimes leads to verbosity in SIL proofs.

\input{latex/July2023/SILIntro.tex}

\input{latex/July2023/STLC.tex}
\input{latex/July2023/STLCBind.tex}
\input{latex/July2023/STLCSafe.tex}

\section{Constructing a Step-Indexed Logic}

Most of the challenge in defining a step-indexed logic is in the
definition of recursive predicates and their fixpoint theorem. The
step-indexed models of \citet{Appel:2001aa} rely on an approximation
operator, here written as $↓_n$, to define the semantic criteria for
valid recursive definitions.  For example, they define a nonexpansive
function to be one that satisfies
\[
  ↓_n f(x) = ↓_n f(↓_n x)
\]
The step-indexed models of \citet{JUNG:2018aa} instead rely on
a step-indexed notion of equality. For example, they define a
nonexpansive function as one that satisfies
\[
  x \stackrel{n}{=} y \text{ implies } f(x) \stackrel{n}{=} f(y)
\]
We are not aware of any inherent reason to prefer one approach to the
other, so we choose the approach of \citet{Appel:2001aa} because they
provide perspicuous proofs of their fixpoint theorem.  So our approach
to SIL can be seen as an implementation of the LSLR logic of
\citet{Dreyer:2011wl} using the foundations of
\citet{Appel:2001aa}. In particular, SIL defines the recursive
predicate of SIL using function iteration, like this:
\[
  ⟦ μ F ⟧ \, x \, n = F^{n\plus 1}\, \top \, x \, n
\]
whereas LSLR uses substitution in the semantics of recursive types:
\[
   ⟦ μ r. R ⟧ \, x \, n = ⟦ R[μr.R/r] ⟧\,x\,n
\]

@@

We organize the rest of the article as follows. We review Appel an
McAllester's work in Section~\ref{sec:appel-mcallester}, then
introduce our representation of step-indexed propositions and
predicates in Agda in Sections \ref{sec:propositions} and
\ref{sec:predicates}.  We formalize several important concepts in
Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
iteration, and prove some basic results about them.  We construct all
of the logical connectives of SIL in
Section~\ref{sec:open-propositions}, including the definition of
recursive predicates and their fixpoint theorem.  The proof theory of
SIL is developed in Section~\ref{sec:proofs}, where we construct the
proof rules of this modal logic.

@@

\input{latex/July2023/StepIndexedLogic.tex}

@@

\input{latex/July2023/STLCDeterministic.tex}
